Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-31 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial economic, social and cultural values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local climate, support the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The simultaneous increase of forests products demand and the substantial climatic changes currently heighten the pressure on remaining forests [@Gibson2011a; @Morales-Hidalgo2015], threatening the maintenance of communities structure, composition and functioning and their dynamics in space and time [@Anderson-Teixeira2013;@Sist2015].

In tropical forest, ecological communities are constantly re-shaped by the natural disturbance events changing both abiotic environment, through the fluxes of light, heat and water, and biotic interactions and competitive pressure [@Goulamoussene2017]. 
The cornerstone of tropical forests ecology is to understand the mechanisms and the determinants of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter consistently modeled and gave important insights into the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009]. 
However the response of forests diversity in tree species remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes following disturbance [@Baraloto2012a].

In the short-term disturbance demonstrated negligible or even positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Still, validations of the IDH remain scarce in the long term and mainly rely on species richness analyses that gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More releveant monitoring would encompass communities composition, crucial for conservation issues, and complete diversity profiles encompassing species evenness in addition to their richness, that reveals ecological rules shaping communities [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, account for species biological attributes and role in the ecosystem is essential to understand the correlations between ecosystems biodiversity, functioning and environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
In that respect functional approaches based on major traits related to species ecology and performance was largely adopted [@Diaz2005;@Villeger2008a], for example highlighting in tropical rainforests the environmental filters fostering fast growing species with efficient resources acquisition after disturbance [@Molino2001;@Haddad2008;@Ruger2009].
Post-disturbance trajectories then corresponded to a shift from "conservative" slow-growing species dealing with scarce resources dominating before disturbance, to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift translated into consistent trajectories for key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction and life history traits (seed mass and maximum height) [@Wright2004;@Westoby2006a;@Chave2009b].

Functional and taxonomic trajectories are both essentiall to fully assess communities response to disturbance.
Both relate to distinct ecological rules and processes and might then be decoupled as it was observed already in tropical forests [@Lohbeck2015; @Guariguata2001].
Although taxonomic and functional diversity are complementary, they are combine into communities functional redundancy that measures the amount of species sharing same trait values [@Carmona2016].
Functional redundancy then explicits the link between otherwise decoupled taxonomic and functional diversities and is central for communities description.
Besides, functional redundancy determines communities resilience as high redundancy, like in highly diverse tropical forests [@Bellwood2006], mitigates the impacts of species removal on ecosystem functioning [@Trenbath1999; @Elmqvist2003; @Diaz2005].

Grasp all facets of communities response to disturbance amount to disentangle the taxonomic and functional trajectories in diversity and composition.
These trajectories would highlight the ecological rules constraining, or not, communities dynamics towards a recovery of initial composition, diversity and functioning.
They would therefore clarify the tenants in the long term of the debated Intermediate Disturbance Hypothesis for tropical forests, which would be as much insights for future adaptive conservation strategies [@Adler2007].
Here we monitored over 30 years the response of 75 ha of forests plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition and diversity. Specifically, we (i) questioned the coupling between taxonomic and functional response to disturbance and identified the underlying assembly processes, which allowed to (ii) clarify the validity of the IDH in the long term for tropical forest and (iii) question the completeness of communities recovery regarding their functional redundancy.

>>>>

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local climate, support the carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The simultaneous increased demand in forests products and the substantial climatic changes currently heighten the pressure on remaining forests [@Gibson2011a; @Morales-Hidalgo2015], threatening the maintenance of communities structure, composition and functioning and their dynamics in space and time [@Anderson-Teixeira2013;@Sist2015].

In tropical forest, ecological communities are constantly re-shaped by natural disturbance events that change both abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and biotic interactions like competition among species [@Herault2018]. 
The cornerstone of tropical forests ecology is to understand the mechanisms and the determinants of ecosystems response to disturbance [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
These structural parameters were thereafter sucessfully modeled, giving important insights into the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009]. 
However the response of forests diversity in tree species remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014, @Liang2016] and would be most probably impacted by the changes following disturbance [@Baraloto2012a].

In the short-term disturbance demonstrated negligible or even positive impacts on communities diversity, which have been formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity at intermediate disturbance intensity [@Molino2001;@Kariuki2006a;@Berry2008a]. 
Still, validations of the IDH remain scarce in the long term and mainly rely on species richness analyses that give limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
More releveant monitoring would encompass communities composition, crucial for conservation issues, and complete diversity profiles integrating both species richness and evenness, to reveal the ecological rules structuring communities [@Magurran1988;@Lavorel2002;@Bellwood2006].
Furthermore, the functional approach that accounts for species biological attributes and role in the ecosystem would be essential to understand the correlations between ecosystems biodiversity, functioning and environmental constraints [@Violle2007b;@Moretti2009;@Baraloto2012a;@Scheiter2013].
The functional trait-based approach focused on major traits related to species ecology and performance was sucessfully adopted [@Diaz2005;@Villeger2008a], for example in highlighting deterministic processes in tropical rainforests that foster fast growing species with efficient resources acquisition after disturbance [@Molino2001;@Haddad2008;@Ruger2009].
Communities then shifted from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@TerSteege2001;@Reich2014;@Herault2011].
This shift translated into consistent trajectories for key functional traits related to resource acquisition (leaf area, density and chlorophyll content, and stem specific gravity and bark thickness), tree growth and reproduction and life history traits (seed mass and maximum height) [@Wright2004;@Westoby2006a;@Chave2009b].

A combination of taxonomic and functional approaches are essential to fully assess communities response to disturbance and differences in their respective trajectories are insightful of underlying mechanisms [@Lohbeck2015;@Guariguata2001].
Although taxonomic and functional diversity are complementary and sometimes decouples, they are combined in the measure of communities functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016].
Functional redundancy is determinant of communities resilience as high redundancy, like in highly diverse tropical forests [@Bellwood2006], mitigates the impacts of species removal on ecosystem functioning [@Trenbath1999; @Elmqvist2003; @Diaz2005].

Grasp all facets of communities response to disturbance comes to examine the taxonomic and functional trajectories in diversity and composition.
These trajectories would highlight the ecological rules constraining or not communities dynamics towards the recovery of initial composition, diversity and functioning.
They would therefore clarify the tenants in the long term of the debated Intermediate Disturbance Hypothesis for tropical forests, a cruicial point for future adaptive conservation strategies [@Adler2007].
Here we monitored over 30 years the response of 75 ha of forests plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed functional traits and species maximum height to draw the trajectories over time of communities taxonomic and functional composition and diversity. Specifically, we (i) questioned the coupling between taxonomic and functional response to disturbance and identified the underlying assembly processes, which allowed to (ii) clarify the validity of the IDH in the long term for tropical forest and (iii) question the completeness of communities recovery regarding their functional redundancy.
